---
title: "Final Model"
author: "Rocky Rowell"
date: "2025-04-17"
output: html_document
---
```{r message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse); library(ggplot2); library(ggthemes); library(lubridate); library(car); library(reshape2); library(caret)
library(mgcv); library(gratia)
```

# Final Model
After spending weeks searching for the best model to forecast Initial Claims in SC, I believe the best model will be using data from 2001 (excludes S&P500) and predict log(InitialClaims Per Capita) with the predictors: Unemployment Per Capita, Labor Force Per Capita, and Layoffs. 

## Data
```{r}
# data is from the CleaningData.rmd file that combines all datasets used into one final dataset
data <- read.csv("Datasets/FinalData.csv", header=T)

# convert date to date format
data$Date <- as.Date(data$Date)

# subset of data starting in 2001
data.2001 <- data %>%
  filter(Date >= "2001-01-01",
         !year(Date) %in% c(2020, 2025),
         Date < "2024-12-31")

# check data
tail(data.2001, 10)
```

## Model Creation
```{r}
# model
final.model <- lm(log(InitialClaims.Lag.PerCapita) ~ Unemployment.PerCapita + LaborForce.PerCapita + Layoffs, data = data.2001)

# check model
summary(final.model)
```


### Diagnostic Plots
```{r}
# plot diagnostic plots
par(mfrow=c(2,2))
plot(final.model)

# check for outliers
par(mfrow=c(1,1))
plot(final.model, 4)
```


### Check Multicollinearity
```{r}
# check vif values
print(vif(final.model))
```


### Check Confidence Intervals
```{r}
# confidence intervals
confint(final.model)
```


## Check Model's Effectiveness at Predicting Initial Claims
### Splitting into Training and Testing Data
```{r}
# set seed
set.seed(112233)

# split the data (70-30)
train.indices <- createDataPartition(data.2001$InitialClaims.Lag.PerCapita, p = 0.7, list = FALSE)

train.data <- data.2001[train.indices, ]
test.data <- data.2001[-train.indices, ]

# create training model
training.model <- lm(log(InitialClaims.Lag.PerCapita) ~ Unemployment.PerCapita + LaborForce.PerCapita + Layoffs, data = train.data)

# create predictions
training.model.pred <- predict(training.model, newdata = test.data)

# MSE and RMSE
MSE <- mean((test.data$InitialClaims.Lag.PerCapita - training.model.pred)^2)
RMSE <- sqrt(MSE)

print(MSE)
print(RMSE)

# plot predictions
ggplot(test.data, aes(x = log(InitialClaims.Lag.PerCapita), y = training.model.pred)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  labs(title = "Predicted vs Actual",
       x = "Actual Values",
       y = "Predicted Values")
```


### Plotting Predicted Initial Claims over Actual Initial Claims
```{r}
# gather predictions
predictions <- predict(final.model, newdata = data.2001)

# fix corrections
predictions <- exp(predictions)

# make new variable in dataset for the predicted intial claims
data.2001$PredictedInitialClaimsLaggedPerCapita <- predictions

# Un-Lag the Predictions
data.2001$PredictedInitialClaimsPerCapita <- lead(data.2001$PredictedInitialClaimsLaggedPerCapita, 1)

# Un-Per Capita the Predictions
data.2001$PredictedInitialClaims <- data.2001$PredictedInitialClaimsPerCapita * data.2001$Population

# check data
print(data.2001, 5)
```

```{r}
# create plot with both actual and predicted initial claims
ggplot(data.2001, aes(x = Date)) +
  # add recession bars
  geom_rect(aes(xmin=as.Date("2001-3-01"), xmax=as.Date("2001-11-01"), ymin=-Inf, ymax=Inf), fill="red2", alpha=0.002) +
  geom_rect(aes(xmin=as.Date("2007-12-01"), xmax=as.Date("2009-06-01"), ymin=-Inf, ymax=Inf), fill="red2", alpha=0.002) +
  # add rest
  geom_line(aes(y = InitialClaims, color = "Actual Initial Claims")) +
  geom_line(aes(y = PredictedInitialClaims, color = "Predicted Initial Claims")) +
  labs(title = "Actual vs Predicted Initial Claims", color = "color",
       x = "Date",
       y = "Initial Claims") +
  scale_color_manual(values = c("Actual Initial Claims" = "steelblue4", "Predicted Initial Claims" = "salmon4")) +
  theme_fivethirtyeight()
```









