---
title: "Markov Chain"
author: "Rocky Rowell"
date: "2025-04-09"
output: html_document
---

```{r message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse); library(scales); library(dplyr)
library(markovchain); library(zoo); library(ggplot2)
library(ggthemes)
```

***
# Comparing US States Unemployment Insurance Claims Transition Probability Matrices
The goal of this project is to compare the transition probability matrices of different states' unemployment insurance claims. 

## Data
The data used in this project is from the US Department of Labor and includes the number of initial claims for each state from 1971 to 2024. The data is transformed into quarterly data in order to account for rapid month to month fluctuations. 





# Data Collection and Preparation
## Load in Claims Data and Fix It
```{r}
# for this file we will have to restart with the raw claims data in order to use multiple states

# load in claims data
claims <- read.csv("Datasets/ClaimsData/ar5159.csv", header = T)

# fix date column
colnames(claims)[colnames(claims) == "rptdate"] <- "Date"
claims$Date <- as.Date(claims$Date, format = "%m/%d/%Y")

# rename state column (personal preference)
claims$State <- as.character(claims$st)

# create InitialClaims column by combining c2-c7
claims <- claims %>%
  mutate(InitialClaims = c2 + c3 + c4 + c5 + c6 + c7) %>%
  select(Date, State, InitialClaims) # drop unnecessary columns

# drop 2020, 2021 and 2025 from dataset to avoid outlier issues
claims <- claims %>%
  filter(!year(Date) %in% c(2020, 2021, 2025))

# make data quarterly to avoid frequent fluctuations
claims.quarterly <- claims %>%
  group_by(State, year_quarter = as.yearqtr(Date)) %>%
  summarise(InitialClaims = sum(InitialClaims, na.rm = TRUE), .groups = "drop") %>% # sum all claims for quarter
  mutate(Date = as.Date(year_quarter, frac = 1)) %>%
  select(State, Date, InitialClaims)
```

## Function to Collect Data
```{r}
# this function will be used to collect the data for a given state

CollectData <- function(state) {
  # filter dataset for the matching state
  data <- claims.quarterly %>%
    filter(State == state)
  
  # return filtered dataset
  return(data)
} 
```

## Function to Graph Initial Claims
```{r}
# this function will be used to graph the initial claims for a given state

GraphClaims <- function(data) {
  ggplot(data, aes(x = Date, y = InitialClaims)) +
    geom_line() +
    scale_y_continuous(labels = comma) +
    labs(title = "Initial Claims over Time (excluding 2020, 2021)", x = " ", y = " ") +
    theme_minimal() + 
    geom_hline(yintercept = mean(data$InitialClaims), linetype="dashed", color = "red")
}
```


***
# Create Markov Chain
## Function to Create Markov Chain
```{r}
# this function will be used to create the markov chain for a given state (data must be filtered first)

CreateMarkovChain <- function(data) {
  # find quantiles
  quantiles <- quantile(data$InitialClaims, probs = c(0.33, 0.66))
  
  # add quantiles to dataset
  data <- data %>%
    mutate(
      InitialClaimsQuantile = case_when(
        InitialClaims <= quantiles[1] ~ "Low",
        InitialClaims > quantiles[1] & InitialClaims <= quantiles[2] ~ "Medium",
        InitialClaims > quantiles[2] ~ "High"
      )
    )
  
  # create vector of states
  states <- data$InitialClaimsQuantile
  
  # fit the markov chain model
  mc_fit <- markovchainFit(data = states)
  mc <- mc_fit$estimate
  
  # vector of correct order (MC doesn't want to do the correct order)
  order <- c("Low", "Medium", "High")
  
  # reorder the matrix
  mc_ordered <- new("markovchain",
                    states = order,
                    transitionMatrix = mc@transitionMatrix[order, order],
                    name = "UI Claims (Ordered)")
  
  # print all results of the markov chain
  print("Summary:")
  summary(mc_ordered)
  print("Transition Matrix:")
  print(mc_ordered)
  print("Steady-State Probabilities:")
  print(steadyStates(mc_ordered))
  
  # plot
  plot(mc_ordered)
}
```


***
# Run Markov Chain on Selected States
## South Carolina
```{r}
sc <- CollectData("SC")
CreateMarkovChain(sc)
```

## North Carolina
```{r}
nc <- CollectData("NC")
CreateMarkovChain(nc)
```

## Georgia
```{r}
ga <- CollectData("GA")
CreateMarkovChain(ga)
```




***
# Is Population an Issue
## South Carolina Claims
```{r}
# plot of sc claims
GraphClaims(sc)
```

## North Carolina Claims
```{r}
# plot of sc claims
GraphClaims(nc)
```

## Georgia Claims
```{r}
# plot of sc claims
GraphClaims(ga)
```






